Computational biology in the 21st century:\\Scaling with compressive algorithms

### 2015-09-29 response.tex

\emph{(2-3) I fully agree that the focus on compressive algorithms should be improved and that title should be more specific.}

We have changed the title, and attempted to narrow the focus to compressive algorithms.

\emph{a. Numerous references are made to [4], sometimes several times within the same paragraph. I realize this was a way to get around the limit on citations, but the result is odd. One solution would be to cite all the original references but split some of them to the supplement.}

We believe we have improved this situation, not by moving references to the supplement, but by pushing this reference to the end when needed, and by citing
the original paper when possible (we freed up some space for references by narrowing the focus of the paper).

@@

Homology search is now defined in the box, and we have improved the language here.

\emph{c. page 5 right, what does ``unblock many blocked reactions'' mean?}

As part of the re-focusing of the paper, we limit discussion of this and other network approaches to a brief mention, and this terminology is no longer present.

\emph{d. page 6 left ``compressive genomics'' is used here and is only defined later. $B_D$ is not defined. The bottom paragraph was vague for me. }

We have defined $B_D$, rewritten the paragraph, and more properly introduced the term ``compressive genomics.''

@@

We have made this change.

\emph{[6] Burroughs-Wheeler is misspelled in section 1. (It should be Burrows)}

We have corrected this error.

\emph{[7] The sample size argument of section 5, paragraph 2 [there are a possible
$20^500$ proteins but only $6.9 x 10^7$ are observed] is not persuasive. Such
``search-space-size'' arguments are nearly always vapid.  This sentence is one of
a over
3984137914278306537107946300187788156651883090392267368064424070371960737746809881430938446547647791637956210590388569173298650466385810245792657795213961405107689148645376
possible strings of characters of this length. But I had to consider very few
of them to write it. I would find a more specific argument about evolution,
relatedness of species, or something else to argue that the low dimensionality
is surprising.}

### 2015-09-29 main.tex
Fix bad bibtex

At the dawn of the genomic era, Sanger sequencing was the most widely-used 
method for reading a genome.
More recently, however, NGS approaches,
beginning with Illumina's `sequencing by synthesis,' have enabled vastly 
greater throughput due to massive parallelism, low cost, and simple sample
preparation.
Illumina sequencing and other NGS approaches
such as SOLiD, Ion Torrent, and 454 pyrosequencing do not read a single DNA
molecule end-to-end as one could read through a bound book.
Instead, in so-called `shotgun sequencing' DNA molecules are chopped into many small fragments;
from these fragments we generate 
\emph{reads} from one or both ends (Figure~\ref{fig:shotgun}).
These reads must be put together in the correct order 
to piece together an entire genome.
Current reads typically range from 50 to 200 bases long, though longer reads are
available with some technologies (e.g., PacBio).
Because no sequencing technology is completely infallible, sequencing machines also
provide a \emph{quality score} (or measure of the confidence in the DNA base
called) associated with each position.
Thus, an NGS read is a string of DNA letters, coupled with a string of ASCII
characters that encode the quality of the base call.
A sequencing run will produce many overlapping reads.

@@

Given a sequenced genome, the next natural questions ask what genes (genomic
regions that code for proteins) are present, what structure each resulting
protein takes, and what biological function it performs.
Identifying likely genes is a well-studied problem~\cite{berger2013computational} 
beyond the scope of this Review.
However, determining evolutionary relationships, structure, and function is
at the heart of current research in computational biology.
Since some organisms (known as \emph{model organisms}) are better studied than
others, and evolution is known to conserve sequence, structure, and function,
a powerful approach to determine these attributes is to search for similar
sequences about which more is known. 
This so-called \emph{homology search} entails searching for \emph{approximate}
matches in databases of known gene or protein sequences.
The homology search problem was previously believed to be solved;
Basic Local Alignment Search Tool (BLAST)~\cite{berger2013computational} has been
the standard tool for performing homology (similarity) search on databases of nucleotide and
protein sequences.
BLAST takes a `seed-and-extend' approach; it looks for small, $k$-mer matches
that might lead to longer matches, and greedily extends them, ultimately 
producing a \emph{sequence alignment} between a query and each potential 
database hit.
However, BLAST's running-time scales linearly with the size of the database being
searched, which is problematic as sequence databases continue to grow at a
faster rate than Moore's law.

On a potentially even larger scale is the growth of metagenomic data.
Metagenomics is the study of the many genomes (bacterial, fungal, and even 
viral) that make up a particular environment.
Such an environment could be soil from a particular region (which can lead to 
the discovery of new 
antibiotics~\cite{forsberg2012shared}), or it could be 
the human gut, whose microbiome has been linked to human-health concerns 
including Autism Spectrum Disorder~\cite{macfabe2012short}, 
Crohn's Disease, and 
obesity.
Metagenomics fundamentally asks what organisms are present, and, in the case
of a microbiome such as the gut, what metabolic functions it can accomplish as
a whole.
One way of addressing this problem is to attempt to map NGS reads from a metagenomic sample
onto a set of reference genomes that are expected to be present.
This is exactly the read-mapping problem discussed above, but with many 
reference genomes, compounding the computational requirements.
A second way is to perform homology search on a protein sequence database;
exact or nearly-exact matches imply the presence of a species, while more 
distant hits may still give clues to function.
For this task, BLASTX~\cite{altschul1990basic} is commonly
used to translate nucleotide reads into their possible protein sequences, and
search for them in a protein database.
The difficulty is that the datasets required to shine any light on these 
questions, namely from `shotgun' metagenomics, are gigantic and vastly more 
complex than standard genomic datasets.
The massive data results in major identification 
challenges for certain bacterial, as well as viral, species and 
genera~\cite{janda200716s}.

The computational study of drugs and their targets based on chemical structure and function
is known as chemogenomics~\cite{bredel2004chemogenomics}.
In the fields of drug discovery and drug repurposing, the prediction 
of biologically active compounds is an important task. 
Computational high-throughput screening eliminates many compounds from 
laborious wet-lab consideration, but even computational screening can be 
time-consuming.
Chemogenomics typically relies on comparing chemical graph structures to identify
similar molecules and binding sites.
Furthermore, comparing chemical graph structures typically involves computing
the maximal common subgraph (MCS), an NP-hard problem.
However, there are an increasing number of such chemical compounds
to search; the NCBI's PubChem database has grown from 31 million compounds in
January 2011 to 68 million in July, 2015.

@@

Compressing reads for storage is necessary should researchers wish to apply more advanced mapping tools or other analysis in the future \cite{bonfield2013compression}.
As stated earlier, NGS reads consist of a sequence string and associated quality scores, the latter of which generally uses more space when compressed.
By taking advantage of biological structure, both parts of NGS reads can be better compressed.
Unlike some other approaches to compressing quality scores in the literature
~\cite{bonfield2013compression,ochoa2013qualcomp},
Quartz~\cite{yu2015quality} takes advantage of the fact that mid-size $l$-mers can in many cases almost uniquely identify locations in the genome,
bounding the likelihood that a quality score is informative and allowing for lossy compression of uninformative scores.
approaches, but slightly improves the accuracy of downstream 
variant-calling~\cite{yu2015quality}.
Because Quartz's lossy compression injects information from the distribution of $l$-mers in the target genome, it demonstrates not only improved compression over competing approaches, but slightly improves the accuracy of downstream variant-calling \cite{yu2015quality}.
Similarly, for the sequence string,
taking advantage of sequence redundancy, Mince~\cite{patro2015data} groups 
similar reads (those that share a common short---15bp---substring) together into `buckets', allowing that common substring to now be 
removed and treated as the bucket label, so that each read in the compressed 
representation comprises only its unique differences from the bucket label.
This approach allows a general-purpose compressor to achieve better compression.
SCALCE~\cite{berger2013computational} also relies on a `boosting' scheme, 
reordering
reads in such a way that a general-purpose compressor achieves improved
compression.

>>>>

At the dawn of the genomic era, Sanger sequencing was the most widely-used 
method for reading a genome.
More recently, however, NGS approaches,
beginning with Illumina's `sequencing by synthesis,' have enabled vastly 
greater throughput due to massive parallelism, low cost, and simple sample
preparation.
Illumina sequencing and other NGS approaches
such as SOLiD, Ion Torrent, and 454 pyrosequencing do not read a single DNA
molecule end-to-end as one could read through a bound book.
Instead, in \emph{shotgun sequencing}, DNA molecules are chopped into many small fragments;
from these fragments we generate 
\emph{reads} from one or both ends (Figure~\ref{fig:shotgun}).
These reads must be put together in the correct order 
to piece together an entire genome.
Current reads typically range from 50 to 200 bases long, though longer reads are
available with some technologies (e.g., PacBio).
Because no sequencing technology is completely infallible, sequencing machines also
provide a \emph{quality score} (or measure of the confidence in the DNA base
called) associated with each position.
Thus, an NGS read is a string of DNA letters, coupled with a string of ASCII
characters that encode the quality of the base call.
A sequencing run will produce many overlapping reads.

@@

Given a sequenced genome, the next natural questions ask what genes (genomic
regions that code for proteins) are present, what structure each resulting
protein takes, and what biological function it performs.
Identifying likely genes is a well-studied problem~\cite{berger2013computational} 
beyond the scope of this Review.
However, determining evolutionary relationships, structure, and function is
at the heart of current research in computational biology.
Since some organisms (known as \emph{model organisms}) are better studied than
others, and evolution is known to conserve sequence, structure, and function,
a powerful approach to determine these attributes is to search for similar
sequences about which more is known. 
This so-called \emph{homology search} entails searching for approximate
matches in databases of known gene or protein sequences.
The homology search problem was previously believed to be solved;
Basic Local Alignment Search Tool (\emph{BLAST})~\cite{berger2013computational} has been
the standard tool for performing homology (similarity) search on databases of nucleotide and
protein sequences.
BLAST takes a `seed-and-extend' approach; it looks for small, $k$-mer matches
that might lead to longer matches, and greedily extends them, ultimately 
producing a sequence alignment between a query and each potential 
database hit.
However, BLAST's running-time scales linearly with the size of the database being
searched, which is problematic as sequence databases continue to grow at a
faster rate than Moore's law.

On a potentially even larger scale is the growth of \emph{metagenomic} data.
Metagenomics is the study of the many genomes (bacterial, fungal, and even 
viral) that make up a particular environment.
Such an environment could be soil from a particular region (which can lead to 
the discovery of new 
antibiotics~\cite{forsberg2012shared}), or it could be 
the human gut, whose microbiome has been linked to human-health concerns 
including Autism Spectrum Disorder~\cite{macfabe2012short}, 
Crohn's Disease, and 
obesity.
Metagenomics fundamentally asks what organisms are present, and, in the case
of a microbiome such as the gut, what metabolic functions it can accomplish as
a whole.
One way of addressing this problem is to attempt to map NGS reads from a metagenomic sample
onto a set of reference genomes that are expected to be present.
This is exactly the read-mapping problem discussed above, but with many 
reference genomes, compounding the computational requirements.
A second way is to perform homology search on a protein sequence database;
exact or nearly-exact matches imply the presence of a species, while more 
distant hits may still give clues to function.
For this task, BLASTX~\cite{altschul1990basic} is commonly
used to translate nucleotide reads into their possible protein sequences, and
search for them in a protein database.
The difficulty is that the datasets required to shine any light on these 
questions, namely from `shotgun' metagenomics, are gigantic and vastly more 
complex than standard genomic datasets.
The massive data results in major identification 
challenges for certain bacterial, as well as viral, species and 
genera~\cite{janda200716s}.

The computational study of drugs and their targets based on chemical structure and function
is known as \emph{chemogenomics}~\cite{bredel2004chemogenomics}.
In the fields of drug discovery and drug repurposing, the prediction 
of biologically active compounds is an important task. 
Computational high-throughput screening eliminates many compounds from 
laborious wet-lab consideration, but even computational screening can be 
time-consuming.
Chemogenomics typically relies on comparing chemical graph structures to identify
similar molecules and binding sites.
Furthermore, comparing chemical graph structures typically involves computing
the maximal common subgraph (MCS), an NP-hard problem.
However, there are an increasing number of such chemical compounds
to search; the NCBI's PubChem database has grown from 31 million compounds in
January 2011 to 68 million in July, 2015.

@@

Compressing reads for storage is necessary should researchers wish to apply more advanced mapping tools or other analysis in the future \cite{bonfield2013compression}.
As stated earlier, NGS reads consist of a sequence string and associated quality scores, the latter of which generally uses more space when compressed.
By taking advantage of biological structure, both parts of NGS reads can be better compressed.
Unlike some other approaches to compressing quality scores in the literature
~\cite{bonfield2013compression,ochoa2013qualcomp},
Quartz~\cite{yu2015quality} takes advantage of the fact that mid-size $l$-mers can in many cases almost uniquely identify locations in the genome,
bounding the likelihood that a quality score is informative and allowing for lossy compression of uninformative scores.
Because Quartz's lossy compression injects information from the distribution of $l$-mers in the target genome, it demonstrates not only improved compression over competing approaches, but slightly improves the accuracy of downstream variant-calling \cite{yu2015quality}.
Similarly, for the sequence component of the read,
Mince~\cite{patro2015data} takes advantage of sequence redundancy by grouping
similar reads (those that share a common short---15bp---substring) together into `buckets,' allowing that common substring to be 
removed and treated as the bucket label, so that each read in the compressed 
representation comprises only its unique differences from the bucket label.
This approach allows a general-purpose compressor to achieve better compression.
SCALCE~\cite{berger2013computational} also relies on a `boosting' scheme, 
reordering
reads in such a way that a general-purpose compressor achieves improved
compression.
